Performance comparison of Dask and Apache Spark on HPC systems \HL{for Neuroimaging}

### 2022-09-12 latex/paper.tex

\section{Introduction}
The rise in data sharing coupled with improved data collection technologies leads neuroimaging into the Big Data era\cite{ALFAROALMAGRO2018400, van2014human, ConpPortal}.
While standard neuroimaging workflow engines, such as Nipype\cite{Nipype:11}, are well rounded to process standard compute-intensive pipelines,
they lack Big Data strategies (i.e., in-memory computing, data locality, and lazy-evaluation) to improve the performance of increasingly prevalent data-intensive pipelines.
A previous study\cite{8752675} noted the importance of in-memory computing and data locality to improve the performance of data-intensive pipelines.
This work studies how performance benchmarks \HL{of neuroimaging applications} generalize across Big Data engines in an HPC environment.
\HL{
	With the rising need for Big Data engines in neuroimaging and none commonly
	used in the field, evaluating the performance of different frameworks is essential.
}
In particular, we study in-depth the performance difference between Dask\cite{Dask:15} and Apache Spark\cite{Spark:16} for their suitability to process neuroimaging pipelines.
\HL{
	Our main contribution is to assess whether Spark or Dask has substantial performance benefits in processing neuroimaging pipelines in HPC systems.
}

@@

\HL{
	While our infrastructure is not of the scale of TOP500 HPC systems, it is 
	hard to reserve multiple nodes concurrently on those infrastructures.
	Moreover, by using our infrastructure, we can ensure that no external
	processes influence the benchmark for our experiments.
	Our system uses similar hardware as Compute Canada's B\'eluga cluster
	--- a representative cluster available to researchers in Canada --- and
	replicates as close as possible its configuration.
	This lets us compare Spark and Dask in an isolated environment with the
	condition of a realistic allocation of resources, for neuroimaging
	applications, in an HPC system.
}

@@

\HL{
	We chose these datasets as they represent typical dataset sizes for neuroimaging studies and their free access.
}

\subsection{Applications}
This section describes the applications used to benchmark Dask and Spark.
We use three simple synthetic applications to have a deep understanding of the 
underlying behavior of the applications with simple I/O patterns, namely
\textit{Increment}, \textit{Multi-Increment}, and \textit{Histogram}.
Also, two neuroimaging applications are used to study more realistic 
applications, namely \textit{Kmeans} and \textit{BIDS App Example}{\cite{gorgolewski2017bids}}.
\HL{
	We chose these neuroimaging data processing applications to study the 
	effects of various Big Data characteristics.
	Table~{\ref{table:applications}} depicts the Big Data characteristics used in
	each application.
}

@@

\HL{
	To minimize the performance difference from implementation decisions, the
	code base of the applications is common for both engines.
	Therefore, the performance observed comes directly from the 
	engine scheduling and APIs.
}

The scripts used for our experiments and the results generated are available on
GitHub at: \mbox{\href{https://github.com/mathdugre/paper-big-data-engines}{https://github.com/mathdugre/paper-big-data-engines/}}.

@@

\subsubsection{Kmeans}
For our third application, we apply Kmeans clustering to the voxel
intensities of the BigBrain image. We set the number of clusters to 3. The
application starts by reading the image blocks, combining all voxels in a
1-D array, and choosing initial centroids using the min, max, and
intermediate values. It assigns each voxel to its closest centroid and
updates each centroid with the average of its assigned voxels. It repeats
the assignment and updates steps for a configurable number of iterations.
Finally, the voxels of the image blocks are classified and written back to
the file system. Updating the centroids involves substantial data
communication between the workers.
The computation to find the closest centroids for each voxels scales with the number of centroids.
It can be computed in parallel at the cost of higher memory or sequentially to reduce memory usage.
We decided to compute the nearest centroids in parallel since we use only three centroids.
However, due to resources limitation on our infrastructure, the Kmeans experiments
are performed with only half of the BigBrain dataset ({\SI{303}{\gibi\byte}}).
\HL{We chose this application for its real-life use case, large data shuffling between workers,
and requirements to keep all data in-memory.}

\subsubsection{BIDS App Example}
Our fourth application is the BIDS App Example, a neuroimaging pipeline to
measure brain volume from MRIs. For this application, we use the CoRR
dataset. The application extracts the brain volume of each participant
then computes the average for each group of participants. Unlike the other
applications, the BIDS App Example is a command-line executed in a Docker image
(bids/example on DockerHub). We converted the Docker image to a Singularity
image for use in HPC environments, using
\href{https://hub.docker.com/r/singularityware/docker2singularity/tags/}{docker2singularity}
\HL{We chose this application for its real-life use case and it is containerized unlike our other applications.}

@@

\section{Results}
\subsection{Increment}
Figure \ref{fig:increment_worker} depicts the total execution time spent in each code segment of the \textit{Increment} application.
The bars are broken down into Overhead, Read, Write, and other application-specific functions.
The mean of each bar is calculated from ten repetitions, with the error bar representing the standard deviation.
As expected, the compute time, from \textit{Increment}, remains the same when we vary the number of nodes.
However, the \textit{Read} and \textit{Write} time increases with the number of nodes for both engines.
\HL{
	With a large number of worker threads accessing the Lustre file system
	concurrently, the disk bandwidth or network is saturated, and it creates 
	an I/O bottleneck to access the Lustre file system.
}

From Figure \ref{fig:increment_worker}, we also observe that Dask's overhead is higher than Spark's.
While the overhead of both engines increases with the number of nodes, Dask's overhead is significantly worse when scaling.
Although Dask has a higher overhead, the total time for both engines is overall similar.
This is due to a tradeoff between disk bandwidth and compute/overhead time.
\HL{
	On the one hand, when multiple threads access Lustre concurrently, they risk 
	saturation of the shared disk bandwidth, thus resulting in longer I/O time.
	On the other hand, when most threads are either idle or performing computations,
	it reduces the contention on the disk bandwidth resulting in shorter I/O time.
}
The only significant difference in total time is when two nodes are used.
This results in an approximately 3\% total runtime difference.
\begin{figure}[!h]
	\centering
	\includegraphics[clip,width=0.75\columnwidth]{figures/stacked_increment_worker.jpg}
	\caption{Increment: varying nodes -- 5000 files, 5 iterations}
	\label{fig:increment_worker}
\end{figure}

@@

In a follow-up experiment, we further optimized our Kmeans implementations for both Dask and Spark to minimize the size of data kept in memory.
Figure~{\ref{fig:dask_kmeans_gantt}} and Figure~{\ref{fig:spark_kmeans_gantt}} show the Gantt chart for the execution
of this new experiments for Dask and Spark.
The x-axis shows the time in second while the y-axis depicts the workers, with 
one worker thread ID (Dask) and process ID (Spark) per line.
We represent each task on the Gantt chart with its associated function's color.
We again observe that Spark reports three times more processes than Dask threads.
\HL{
	It is unclear what the cause is for reporting extra Spark processes.
	However, with our analysis, we found that both Dask and Spark use the same 
	amount of maximum concurrent processes or threads.
	Additionally, the is no significant difference in the average amount of 
	concurrent processes or threads between Dask and Spark.
}

@@

\HL{
	Another study~{\cite{Slaughter2020-ql}} developed a parametrized benchmark framework
	to evaluate the performance of different engines, including Spark and Dask.
	On the one hand, the framework reduces the implementation effort and helps with
	comparing frameworks fairly.
	On the other hand, it is hard to characterize neuroimaging realistically
	applications using their framework since the runtime of these applications can vary
	greatly depending on the subject process, and the framework lacks kernels with
	a variable amount of compute-intensity.
	Due to these limitations, we decided to implement the benchmarked applications
	directly into Spark and Dask.
}

\subsection{Recommendations}
\HL{
	The results from our experiments show that the performance difference between
	Dask and Spark are minimal, in general. 
	The main differences we observed are qualitative.
	First, we found that Dask integration with the scientific Python ecosystem made it
	easier to implement neuroimaging applications.
	Spark configuration requires setting multiple environment variables.
	In comparison, Dask is ready to go after an installation through PIP.
	In addition, Dask dashboard provides more information than Spark ones, making
	the debugging of Dask applications easier and faster.
	Therefore, for the neuroimaging field, we recommend using Dask over Spark.
	However, in other fields with non-Python-based ecosystems, both Dask and Spark
	would be equal candidates as Big Data engines.
}

>>>>

\section{Introduction}
The rise in data sharing coupled with improved data collection technologies leads neuroimaging into the Big Data era\cite{ALFAROALMAGRO2018400, van2014human, ConpPortal}.
While standard neuroimaging workflow engines, such as Nipype\cite{Nipype:11}, are well rounded to process standard compute-intensive pipelines,
they lack Big Data strategies (i.e., in-memory computing, data locality, and lazy-evaluation) to improve the performance of increasingly prevalent data-intensive pipelines.
A previous study\cite{8752675} noted the importance of in-memory computing and data locality to improve the performance of data-intensive pipelines.
This work studies how performance benchmarks \HL{of neuroimaging applications} generalize across Big Data engines in an HPC environment.
\HL{
	With the rising need for Big Data engines in neuroimaging and none commonly
	used in the field, evaluating the performance of different frameworks is essential.
}
In particular, we study in-depth the performance difference between Dask\cite{Dask:15} and Apache Spark\cite{Spark:16} for their suitability to process neuroimaging pipelines.
\HL{
	Our main contribution is to assess whether Spark or Dask has any substantial performance advantage in processing neuroimaging pipelines in HPC systems.
}

@@

\HL{
	While our infrastructure is not of the scale of TOP500 HPC systems,
	we can assure that no external processes
	impacted our measurements which would be hard to achieve on a production system.
	Our system uses similar hardware as Compute Canada's B\'eluga cluster
	--- a representative cluster available to researchers in Canada --- and 
	replicates as closely as possible its configuration.
	This lets us compare Spark and Dask in an isolated environement with the condition 
	of a realistic  allocation of resources, for neuroimaging applications, in an HPC system.
}

@@

\HL{
	We chose these dataset as they represent common dataset sizes for neuroimaging 
	imaging studies and are open-access.
}

\subsection{Applications}
This section describes the applications used to benchmark Dask and Spark.
We use three simple synthetic applications to have a deep understanding of the 
underlying behavior of the applications with simple I/O patterns, namely
\textit{Increment}, \textit{Multi-Increment}, and \textit{Histogram}.
Also, two neuroimaging applications are used to study more realistic 
applications, namely \textit{Kmeans} and \textit{BIDS App Example}{\cite{gorgolewski2017bids}}.
\HL{
	We chose these neuroimaging data processing applications to study the 
	effects of various Big Data characteristics.
	Table~{\ref{table:applications}} depicts the Big Data characteristics used in
	each application. 
}
\TG{Explain what each column represents}

@@

\HL{
	To minimize the performance difference from implementation decisions, the
	code base of the applications is common for both engines.
	Therefore, the performance observed comes directly from the engine scheduling and
	their APIs.
} The scripts used for our experiments and the results generated are available on
GitHub at: \mbox{\href{https://github.com/mathdugre/paper-big-data-engines}{https://github.com/mathdugre/paper-big-data-engines/}}.

@@

\subsubsection{Kmeans}
For our third application, we apply Kmeans clustering to the voxel
intensities of the BigBrain image. We set the number of clusters to 3. The
application starts by reading the image blocks, combining all voxels in a
1-D array, and choosing initial centroids using the min, max, and
intermediate values. It assigns each voxel to its closest centroid and
updates each centroid with the average of its assigned voxels. It repeats
the assignment and updates steps for a configurable number of iterations.
Finally, the voxels of the image blocks are classified and written back to
the file system. Updating the centroids involves substantial data
communication between the workers.
The computation to find the closest centroids for each voxels scales with the number of centroids.
It can be computed in parallel at the cost of higher memory or sequentially to reduce memory usage.
We decided to compute the nearest centroids in parallel since we use only three centroids.
However, due to resources limitation on our infrastructure, the Kmeans experiments
are performed with only half of the BigBrain dataset ({\SI{303}{\gibi\byte}}).
\HL{We chose this application as it involves substantial data shuffling between workers,
and requires to keep all the data in memory.}

\subsubsection{BIDS App Example}
Our fourth application is the BIDS App Example, a neuroimaging pipeline to
measure brain volume from MRIs. For this application, we use the CoRR
dataset. The application extracts the brain volume of each participant
then computes the average for each group of participants. Unlike the other
applications, the BIDS App Example is a command-line executed in a Docker image
(bids/example on DockerHub). We converted the Docker image to a Singularity
image for use in HPC environments, using
\href{https://hub.docker.com/r/singularityware/docker2singularity/tags/}{docker2singularity}
\HL{We chose this application for its real-life use case and the fact that it is containerized unlike our other applications.}

@@

\section{Results}
\subsection{Increment}
Figure \ref{fig:increment_worker} depicts the total execution time spent in each code segment of the \textit{Increment} application.
The bars are broken down into Overhead, Read, Write, and other application-specific functions.
The mean of each bar is calculated from ten repetitions, with the error bar representing the standard deviation.
As expected, the compute time, from \textit{Increment}, remains the same when we vary the number of nodes.
However, the \textit{Read} and \textit{Write} time increases with the number of nodes for both engines.
\HL{
	Due to the large amount of worker threads accessing the Lustre file system
	concurrently, the disk bandiwdth or network are saturated, which results in an I/O bottleneck.
}

From Figure \ref{fig:increment_worker}, we also observe that Dask's overhead is higher than Spark's.
While the overhead of both engines increases with the number of nodes, Dask's overhead is significantly worse when scaling.
Although Dask has a higher overhead, the total time for both engines is overall similar.
This is due to a tradeoff between disk bandwidth and compute/overhead time.
\HL{
	On the one hand, when multiple threads access Lustre concurrently they may saturate the shared disk bandwidth, resulting in longer I/O time.
	On the other hand, when most threads are either idle or performing computations,
	the contention on the disk bandwidth is reduced, resulting in shorter I/O time.
}
The only significant difference in total time is when two nodes are used.
This results in an approximately 3\% total runtime difference.
\begin{figure}[!h]
	\centering
	\includegraphics[clip,width=0.75\columnwidth]{figures/stacked_increment_worker.jpg}
	\caption{Increment: varying nodes -- 5000 files, 5 iterations}
	\label{fig:increment_worker}
\end{figure}

@@

In a follow-up experiment, we further optimized our Kmeans implementations for both Dask and Spark to minimize the size of data kept in memory.
Figure~{\ref{fig:dask_kmeans_gantt}} and Figure~{\ref{fig:spark_kmeans_gantt}} show the Gantt chart for the execution
of this new experiments for Dask and Spark.
The x-axis shows the time in second while the y-axis depicts the workers, with 
one worker thread ID (Dask) and process ID (Spark) per line.
We represent each task on the Gantt chart with its associated function's color.
We again observe that Spark reports three times more processes than Dask threads.
\HL{
	It is unclear what is the cause for reporting extra Spark processes.
	However, we found that both Dask and Spark use the same 
	amount of maximum concurrent processes or threads.
	Additionally, there was no significant difference in the average amount of 
	concurrent processes or threads between Dask and Spark.
}

@@

\HL{
	Another study~{\cite{9355308}} developed a parametrized benchmark framework
	to evaluate the performance of different engines, including Spark and Dask.
	On the one hand, the framework reduces the implementation effort and helps with
	comparing frameworks fairly.
	On the other hand, it is hard to characterize neuroimaging realistically
	applications using their framework since the runtime of these applications can vary
	greatly depending on the subject process, and the framework lacks kernels with
	a variable amount of compute-intensity.
	Due to these limitations, we decided to implement the benchmarked applications
	directly into Spark and Dask.
}

\subsection{Recommendations}
\HL{
	The results from our experiments show that the performance difference between
	Dask and Spark are minimal, in general. 
	The main differences we observed are qualitative.
	We found that Dask integration with the scientific Python ecosystem made it
	easier to implement neuroimaging applications.
	Spark configuration required to set multiple environment variables.
	In comparison, Dask is ready-to-go after an installation through PIP, the Python package manager.
	The Dask dashboard provides more information than Spark's, making
	the debugging of Dask applications easier and faster.
	Therefore, for the neuroimaging field, we recommend the use of Dask over Spark.
	However, in other fields with non-Python-based ecosystem, both Dask and Spark
	may be equal candidates as a Big Data engine.
}
